---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tpdmethods

<!-- badges: start -->
<!-- badges: end -->

The goal of tpdmethods is to make TPD-based methods accessible to more researchers. 

Recent work in multivariate extremes is built on the so-called tail pairwise 
dependence (TPD). This summary metric of dependence in the pairwise tail shares 
properties with covariance and this recent work has used these properties to 
construct extremal analogues to classical statistical tools including PCA and 
ARMA time series models. We implement these extremal analysis pipelines in this 
R package 'tpdmethods'. The package includes datasets and vignettes for 
precipitation and fire weather analyses. 

## Installation

You can install the development version of tpdmethods from [GitHub](https://github.com/) with:

``` r
# install.packages("pak") # pak is a method to download packages from a variety of sources.
pak::pak("twixson/tpdmethods")
```

## Example

This is a basic example which shows a simple TLETS workflow. We will perform a 
simplified version of the analysis in Wixson and Cooley (2023). The scientific 
question is how much more likely is a fire season like 2020 under present 
climate than under past climate. Wixson and Cooley (2023) used Fire Weather Index
(FWI) data to answer this question. Their marginally-transformed data lives in 
the package as `fire_weather_present`, and `fire_weather_past`. Each dataset is 
a matrix with 153 rows for the days in the fire season and 20 columns for the 
years.

The first step is to estimate the TPD of the time series data. Since 
these data have plausibly iid replicates (each season) and are already on 
Fr\'echet(2) margins we will use `matrix_as_seasons = TRUE` and 
`trans_marginal = FALSE` in our `tpd()` call. 

```{r 'estimate tpd', fig.height=3, fig.width=7}
library(tpdmethods)
par(mar = c(4, 4, 2, 1) + 0.1)
plot(fire_weather_present[,15], 
     type = "l",
     main = "2020 Season FWI on RV Margins", 
     xlab = "day", 
     ylab = "FWI")

fw_tpd_present <- tpd(fire_weather_present, 
                      max_lag = 30, 
                      trans_marginal = FALSE, 
                      matrix_as_seasons = TRUE)

plot(0:30, fw_tpd_present, 
     type = "h", 
     ylim = c(0, 1), 
     xlab = "lag", 
     ylab = "TPD")
```

After estimating the TPD we need to fit a model to the data. We will use the 
extremes analogue to the innovations algorithm to fit a transformed-linear 
moving average (TL-MA) model up to order `q = 20`. We will visually compare the 
plot of the fitted TPD function to the empirical TPD function to determine which
model to retain. 

```{r 'fit MAs', fig.height=3, fig.width=7}
set.seed(1982374)
model_coefs_present <- innovations(fw_tpd_present, max_q = 20)

par(mar = c(4, 4, 2, 1) + 0.1)
plot(0:30, fw_tpd_present, type = "h", ylim = c(0, 1), 
     main = "Present TPDFs", 
     xlab = "lag", 
     ylab = "TPD value")
lines(1:30 + 0.2, maq_tpdf(model_coefs_present[[1]][15,], max_lag = 30), 
      type = "h", col = "4")
legend("topright", legend = c("empirical", "MA(15)"), 
       text.col = c(1,4), bty = "n")
abline(h = 0.1, lty = 2, col = "grey")
```

Here we can see the TL-MA(15) model fits well up to lag-15. TL-MA(q) models have
non-zero TPDFs only up to lag-q just like classical MA(q) models. Included on 
the plot is a horizontal dashed line at 0.1. This is included because of the 
known bias in the TPD estimator. Wixson and Cooley (2023) considered TPD values 
that were consistently around this line to be that known bias and thus 
continued their analysis with the TL-MA(15). 

We will repeat the previous steps with the past-climate data (from 1958-1979). 

```{r, 'past climate', fig.height=3, fig.width=7}
fw_tpd_past <- tpd(fire_weather_past, 
                   max_lag = 30, 
                   trans_marginal = FALSE, 
                   matrix_as_seasons = TRUE)

model_coefs_past <- innovations(fw_tpd_past, max_q = 20)

par(mar = c(4, 4, 2, 1) + 0.1)
plot(0:30, fw_tpd_past, type = "h", ylim = c(0, 1), 
     main = "Past TPDFs", 
     xlab = "lag", 
     ylab = "TPD value")
lines(1:30 + 0.2, maq_tpdf(model_coefs_past[[1]][15,], max_lag = 30), 
      type = "h", col = "4")
legend("topright", legend = c("empirical", "MA(15)"), 
       text.col = c(1,4), bty = "n")
abline(h = 0.1, lty = 2, col = "grey")
```

Wixson and Cooley (2023) considered a high quantile of the present climate FWI
and considered a season as extreme as 2020 if it had at least as many days above
that high quantile as 2020 did. Here we use the marginally transformed data to 
perform the same analysis for one quantile (the 0.98 quantile). This analysis 
differs from Wixson and Cooley because the marginal transformation removed 
seasonality which included a location-parameter based climate signal. Here we 
are only looking at the dependence information. 

```{r 'get high quantile'}
q098 <- stats::quantile(fire_weather_present, probs = 0.98)
(count_2020 <- sum(fire_weather_present[,15] > q098))
```

We note that a generated TL-MA(q) time series has an unknown marginal 
distribution. The shape of the tail is still 2 but the distribution has changed. 
This means that we will need to marginally transform the generated
seasons back to being marginally Fr\'echet(2). To do this we generate 1000 
seasons all at once and use the `transform_marginal()` function.  

```{r 'generate seasons', cache = TRUE}
set.seed(2389098)
temp_past <- gen_maq(n = 153*1000, thetas = model_coefs_past[[1]][15,1:15])
temp_present <- gen_maq(n = 153*1000, 
                        thetas = model_coefs_present[[1]][15,1:15])
```

```{r}
seasons_past <- 
  matrix(transform_marginal(temp_past, fix_gpd_params = TRUE, gpd_shape = 0.5), 
         ncol = 1000)
seasons_present <- 
  matrix(transform_marginal(temp_present, fix_gpd_params = TRUE, gpd_shape = 0.5), 
         ncol = 1000)

```

Finally, we count the number of seasons with at least `r count_2020` days above
the high quantile. 

```{r 'attribution'}
extreme_past    <- 
  apply(seasons_past, 2, function(x){sum(x > q098) > count_2020})
extreme_present <- 
  apply(seasons_present, 2, function(x){sum(x > q098) > count_2020})

result <- data.frame(quantile = 0.98,
                     threshold = unname(q098),
                     Num_2020 = count_2020,
                     past = sum(extreme_past),
                     present = sum(extreme_present),
                     ratio = sum(extreme_present)/sum(extreme_past))
result
```

It appears that dependence may have been stronger under the climate
of fifty years ago. That hypothesis would need more careful testing, but now you 
see some of the TLETS aspects of `tpdmethods`. 
