---
title: "financial_PCA"
output: rmarkdown::html_vignette
author: Troy P. Wixson & Daniel Cooley
vignette: >
  %\VignetteIndexEntry{financial_PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tpdmethods)
```


## Vignette 2: Principal Components Analysis - Financial Data

In this brief example we will explore the extremes analogue to principal
components analysis using financial data from the Kenneth R. French Data Library.
These data are included in the package as `financial_data`. The data consist of
13599 observations of losses (negative value-averaged daily returns) from 30 
financial sectors (e.g., coal, autos, and retail) from 1970 through 2023. 
Definitions of the financial sectors can be found in `Industry_definitions.txt`. 

### Data Exploration

The first step of the analysis pipeline is to determine whether the data are 
heavy tailed, if they can be reasonably assumed to have the same index of 
regular variation (alpha), and to estimate that alpha. This is required because
the `tpd()` function assumes a heavy tail and a shared alpha. 

To check the tail of these data we will use the Hill estimator (Hill, 1975). 
This estimator uses the largest `k` order statistics to estimate alpha. A 
standard way of choosing `k` is to look at a Hill plot which computes the Hill 
estimator across a range of values for `k` and choosing the largest `k` such 
that the estimates for all smaller values of `k` are relatively stable. We use 
the function `hill()` from the `evir` package on a few variable here to 
illustrate the method. We consider using up to `k = 2000` (the 0.85 quantile) as 
a larger `k` seems implausible. In addition we and add horizontal 
lines to the Hill plots as a visual aid. The first three plots include a line 
at the joint estimate and the last plot includes a second line at a more 
believable value. 

```{r 'hill_plots1', fig.height=7, fig.width=7}
par(mfrow = c(2,2))
  evir::hill(financial_data[,2], end = 2000)
  abline(h = 3.139, col = 4)
  evir::hill(financial_data[,3], end = 2000)
  abline(h = 3.139, col = 4)
  evir::hill(financial_data[,29], end = 2000)
  abline(h = 3.139, col = 4)
  evir::hill(financial_data[,30], end = 2000)
  abline(h = c(2.5, 3.139), col = 4)
par(mfrow = c(1,1))
```

Most of the plots for the 30 financial sectors look similar to the two plots 
in the top row. The plot in the bottom left is hard to interpret as the 
estimates do not appear to stabilize. The plot on the bottom right complicates 
the interpretation as the estimates appear to stabilize but at a different value 
for alpha. In all four plots the estimates for alpha become noisy when `k` is 
quite small which is expected as the variance is greater when there are fewer 
observations used. 

It is challenging to select a single value above which the estimates 
become stable but the horizontal line in these plots suggests that somewhere 
between 300 and 500 order statistics is reasonable. We will move forward with 
`k = 300` because of the Hill plot on the bottom left which suggests a smaller 
number is needed. Selecting `k = 300` corresponds to assuming stability above 
the 0.98 quantile which is high enough be be reasonable. We assume 
that the same `k` is reasonable for all variables. This will not always be the 
case (for example weather station data with different length records would 
likely use different `k`'s). This package cannot handle differing `k`'s. 

Our next step is to determine if it is plausible that the shape of the tail for
all of the variables is the same (i.e., if they share an alpha). To do this we 
use the `alpha_plot()` function which plots the point estimate and confidence 
interval for each variable as well as a horizontal line at the joint estimate. 

```{r "alpha_plot", fig.height=2.5, fig.width=7}
alpha_plot(financial_data[,-1], k = 300)
```

There is broad agreement across financial sectors with two notable exceptions. 
The financial and textiles sectors appear to have smaller alphas (heavier tails)
than the rest of the market. Here the researcher has to make a choice about 
whether a common alpha makes sense or not. There are multiple testing issues 
involved with comparing 30 confidence intervals and that the choice
of `k` has a large effect on the resulting confidence intervals. 

From here a practitioner needs to determine if they want to transform the
margins. If the practitioner wants to transform the margins (as we do in the 
first of our two following analyses) then the assumption of a common alpha is 
unnecessary as the marginal transformation will put all variables on 
Fr\'echet($\alpha = 2$) margins. The practitioner can move forward with the 
marginal transformation whether or not the underlying variables have the same 
alpha. If, instead, the practitioner wants to leave the variables on the 
natural margins then they must be willing to assume a common alpha (as we do in 
the second of the following analyses). If the assumption of a common alpha is
reasonable for some, but not all, of the desired variables then the practitioner 
could drop variables with different tails from the study or transform the 
margins of the few that don't match. 

### PCA Example 1: Correlation Analogue

Transforming the margins can be done to ensure a shared alpha. The method of 
transforming the margins that is built into the `tpdmethods` package assumes 
that a generalized Pareto distribution is a reasonable model above some high 
quantile (the emperical 0.975-quantile by default) and uses a rank 
transformation (the ECDF) below that quantile. This piece-wise model is used to 
put each variable onto uniform margins which are then transformed with the 
probability integral transform to Fr\'echet($\alpha = 2$) margins. This 
transformation, coded in the `transform_marginal()` function, ensures a common 
alpha as well as a common scale. Since the variables now each have unit scale 
the subsequent analysis is analogous to PCA on the correlation matrix rather 
than the covariance matrix. 

To move forward with our analysis we must estimate the TPD matrix (TPDM). The 
TPDM is an extremes analogue to the covariance matrix. It shares many 
properties with a covariance matrix including having a correlation analogue as
a special case. The `tpd()` function is used to estimate the TPDM and it 
includes the marginal transformation by default (`trans_marginal = TRUE`). Here 
we have to exclude the first column as it is the dates of each observation. 

```{r 'pca_tpd'}
tpdm <- tpd(as.matrix(financial_data[,-1]))
```

Now that the TPDM is estimated we perform an eigen decomposition just like in 
classical PCA. The elements of the TPDM are estimated pairwise and thus the 
estimated TPDM may not be positive definite. We fix this by performing the 
eigen decomposition on a positive definite matrix that is close to the 
estimated TPDM using the `make_pd = TRUE` toggle in the `get_eigen()` call. 

```{r 'eigen_decomp'}
eigen_tpdm <- get_eigen(tpdm, make_pd = TRUE)
```

Now that we have the eigen decomposition of the TPDM we can try to interpret 
the eigenvectors and eigenvalues using some plots. Our first plot is a scree 
plot which demonstrates how important the first eigenvalue is. We include this 
plot twice to show that the `normalize` argument only changes the scale. 

```{r 'scree', fig.height=2.5, fig.width=7}
p1 <- plot_scree(eigen_tpdm$values, max_ind = 15, normalize = TRUE)
p2 <- plot_scree(eigen_tpdm$values, max_ind = 15, normalize = FALSE)
cowplot::plot_grid(p1, p2, ncol = 2)
```

The next plot looks at the eigenvector loadings.

```{r 'eigen_plot', fig.height=2.5, fig.width=7}
plot_eigen(eigen_tpdm$vectors, 
           num_vecs = 5, 
           var_names = colnames(financial_data[-1]))
```

This plot shows that the first (and by-far most important) eigenvector
is nearly constant which suggests that large losses in any sector are frequently 
accompanied by large losses across the entire market. The next eigenvalues
distinguish between sub-components of the financial market. The second
eigenvector suggests that the beer, food, health, household, and smoke sectors 
of the market tend to have less extreme losses than more discretionary items 
like autos, books, etc. 


In addition to plots of the eigenvector weights, we can make time series plots 
for a few principal components. 

```{r 'time_series_plots', fig.height=6, fig.width=7}
dates <- as.Date(as.character(financial_data[,1]), "%Y%m%d")
ts1_plot <- get_ts_plot(eigen_num = 1, 
                        eigen_vecs = eigen_tpdm$vectors, 
                        data = as.matrix(financial_data[,-1]), 
                        dates = dates)
ts2_plot <- get_ts_plot(eigen_num = 2, 
                        eigen_vecs = eigen_tpdm$vectors, 
                        data = as.matrix(financial_data[,-1]), 
                        dates = dates)
ts3_plot <- get_ts_plot(eigen_num = 3, 
                        eigen_vecs = eigen_tpdm$vectors, 
                        data = as.matrix(financial_data[,-1]), 
                        dates = dates)
                  
cowplot::plot_grid(ts1_plot, ts2_plot, ts3_plot, nrow = 3)
```

There is a very large negative value in principal component 1. What day is that?

```{r 'investigate_ts'}
ts1 <- get_pc_ts(eigen_num = 1, 
                 eigen_vecs = eigen_tpdm$vectors, 
                 data = as.matrix(financial_data[,-1]))
ts1_ordered <- order(ts1)
dates[head(ts1_ordered)]
```

The largest loss is Black Monday, October 19, 1987, which is the largest 
single-day U.S. stock market loss. The second largest loss, March 16, 2020, is 
the second Black Monday of the 2020 stock market crash due to COVID-19.

Finally, we can look at scatter plots of the first principal components. 
```{r 'scatter', fig.height=2, fig.width=7}
scatter_df <- data.frame(pc1 = ts1, 
                         pc2 = get_pc_ts(eigen_num = 2, 
                                         eigen_vecs = eigen_tpdm$vectors, 
                                         data = as.matrix(financial_data[,-1])))
max_val <- max(abs(scatter_df))

radii <- apply(scatter_df, 1, function(x){sqrt(sum(x^2))})
scatter_df$large <- as.factor(
  ifelse(radii > quantile(probs = 0.95, radii), 1, 0))

ggplot2::ggplot(scatter_df, ggplot2::aes(x = pc1, y = pc2)) + 
  ggplot2::geom_point(ggplot2::aes(col = large), alpha = 0.5) + 
  ggplot2::coord_fixed() + 
  ggplot2::scale_color_discrete(type = scico::scico(4, palette = "vik")[2:3]) + 
  ggplot2::theme_minimal()
```



### PCA Example 2: Covariance Analogue

If we believe that the data are tail-equivalent (i.e., they share an alpha, 
which seems plausible given the plot produced by `alpha_plot()` earlier) then we 
can perform a similar PCA for the covariance analogue (rather than the 
correlation analogue) by setting `trans_marginal = FALSE`. If we want to ensure 
that the alpha associated with the estimated TPDM is the same as what we saw in 
the plot above then we can set `fix_alpha = 3.139` or set `k = 300`. 

The default settings for computing the TPDM computes the radial components of the 
pseudo-polar decomposition pairwise. This is sensible if
we expect the most extreme events to be local (e.g., rainfall). When we expect
large events to occur globally it makes sense to compute the radii as a norm 
across the whole vector. This is done with `vector_norm = TRUE`. 

```{r "cov_analogue", fig.height=2.5, fig.width=7}
tpdm_cov <- tpd(as.matrix(financial_data[,-1]), 
                trans_marginal = FALSE, 
                vector_norm = TRUE)
eigen_cov <- get_eigen(tpdm_cov, make_pd = TRUE)
plot_eigen(eigen_cov$vectors, var_names = colnames(financial_data[-1]))
```

Here we highlight that, just as in classical PCA, when PCA is done on the 
covariance matrix and the scales are different then one of the early 
eigenvectors has to account for that difference in scale. A primary difference 
between these two sets of eigenvectors is the weighting on the coal sector. The 
scales in the correlation-analogue are fixed at one which obscures the large 
scale in coal on the original margins. 

```{r "scale_differences"}
# correlation analogue
head(sort(diag(tpdm), decreasing = TRUE))

# covariance analogue
head(sort(diag(tpdm_cov), decreasing = TRUE))
colnames(financial_data[-1])[order(diag(tpdm_cov))[30:25]]
```

In the preceding example we considered a covariance analogue by leaving the 
variables on their original margins. An alternative method would be to estimate
the alpha for each variable with, for example, the Hill estimator and then to 
perform a power transformation so that all transformed variables have the same 
alpha. A power transformation would retain scale information which would allow
a covariance analogue to the TPDM. This is not currently implemented in the 
`tpdmethods` package. Please reach out if you feel it, or anything else, would 
be a usefule addition. 

### Summary

In this brief example we demonstrated a few of the key PCA analogue functions: 

* `alpha_plot()` - for checking the assumption of a common alpha.
* `tpd()` - for estimating the TPD matrix of in a multivariate setting.
* `get_eigen()` - for performing the eigen decomposition of a positive definite
    matrix that is close to the estimated TPDM
* `plot_scree()` - for creating a scree plot of the eigenvalues. 
* `plot_eigen()` - for creating a plot which shows eigenvector loadings.  
* `get_ts_plot()` - for creating a time series plot of a principal component. 
* `get_pc_ts()` - for getting the principal component time series for further
    investigation. 
