---
title: "financial_PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{financial_PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tpdmethods)
```


## Vignette 2: Principal Components Analysis - Financial Data

In this second breif example we will explore the extremes analogue to principal
components analysis using financial data from the Kenneth R. French Data Library.
These data are included in the package as `financial_data`. The data consist of
13599 observations of losses (negative value-averaged daily returns) from 30 
financial sectors (e.g., coal, autos, and retail) from 1970 through 2023. 
Definitions of the financial sectors can be found in `Industry_definitions.txt`. 

The first step of the analysis pipeline is to determine whether the data are 
heavy tailed, if they can be reasonably assumed to have the same index of 
regular variation (alpha), and to estimate that alpha. This is required because
the `tpd()` function assumes a heavy tail and a shared alpha. 

To check the tail of these data we will use the Hill estimator (Hill, 1975). 
This estimator uses the largest `k` order statistics to estimate alpha. A 
standard way of choosing `k` is to look at a Hill plot which computes the Hill 
estimator across a range of values for `k` and choosing the largest `k` such 
that the estimates for all smaller values of `k` are relatively stable. We use 
the function `hill()` from the `evir` package on a few variable here to 
illustrate the method. We consider using up to k = 2000 and add horizontal 
lines to the Hill plots as a visual aid. 

```{r 'hill_plots1', fig.height=7, fig.width=7}
par(mfrow = c(2,2))
  evir::hill(financial_data[,2], end = 2000)
  abline(h = 3.25, col = 4)
  evir::hill(financial_data[,3], end = 2000)
  abline(h = 3.25, col = 4)
  evir::hill(financial_data[,29], end = 2000)
  abline(h = 3.25, col = 4)
  evir::hill(financial_data[,30], end = 2000)
  abline(h = c(2.5, 3.25), col = 4)
par(mfrow = c(1,1))
```


Notice that the estimate becomes noisy when `k` is quite small as is standard 
with these types of plots. It is challenging to determine where the estimates 
become stable but the horizontal line in these plots suggests that somewhere 
between 300 and 500 is reasonable. We will move forward with `k = 300` because
the Hill plot on the bottom left suggests a smaller number is needed. We assume 
that the same `k` is reasonable for all variables. This will not always be the 
case (for example weather station data with different length records would 
likely use different `k`'s). This package cannot handle differing `k`'s. 

Our next step is to determine if it is plausible that the shape of the tail for
all of the variables is the same (i.e., if they share an alpha). To do this we 
use the `alpha_plot()` function which plots the point estimate and confidence 
interval for each variable as well as a horizontal line at the joint estimate. 

```{r "alpha_plot", fig.height=2.5, fig.width=7}
alpha_plot(financial_data[,-1], k = 300)
```

There is broad agreement across financial sectors with two notable exceptions. 
The financial and textiles sectors appear to have smaller alphas (heavier tails)
than the rest of the market. Here the researcher has to make a choice about 
whether a common alpha makes sense or not. There are multiple testing issues 
involved with comparing 30 confidence intervals and that the choice
of `k` has a large effect on the resulting confidence intervals. If the 
researcher is willing to assume a common alpha then they can move on (as we do 
a bit further below). If not then they can either drop those variables, 
perform a marginal transformation on those variables, or perform a marginal 
transformation on all of the variables in the dataset (as we do here). 
Transforming the margins can be done to ensure a shared alpha and to ensure a 
common scale, this allows the analysis to be analogous to PCA on the correlation
matrix rather than the covariance matrix. 

Our next step is estimating the TPD matrix (TPDM). The TPDM is an extremes 
analogue to the covariance matrix and it shares many 
properties with a covariance matrix. In the following example we have 
standardized the margins so the TPDM is analogous to a correlation matrix. 

```{r 'pca_tpd'}
tpdm <- tpd(as.matrix(-financial_data[,-1]))
```

Now that the TPDM is estimated we need to perform the eigen decomposition. 
The elements of the TPDM are estimated pairwise and thus the estimated TPDM may
not be positive definite. We fix this by performing the eigen decomposition on 
a positive definite matrix that is close to the estimated TPDM using the 
`make_pd = TRUE` toggle in the `get_eigen()` call. 

```{r 'eigen_decomp'}
eigen_tpdm <- get_eigen(tpdm, make_pd = TRUE)
```

Now that we have the eigen decomposition of the TPDM we can try to interpret 
the eigenvectors and eigenvalues using some plots. Our first plot is a scree 
plot which demonstrates how important the first eigenvalue is. We include this 
plot twice to show that the `normalize` argument only changes the scale. 

```{r 'scree', fig.height=2.5, fig.width=7}
p1 <- plot_scree(eigen_tpdm$values, max_ind = 15, normalize = TRUE)
p2 <- plot_scree(eigen_tpdm$values, max_ind = 15, normalize = FALSE)
cowplot::plot_grid(p1, p2, ncol = 2)
```

The next plot looks at the eigenvector loadings.

```{r 'eigen_plot', fig.height=2.5, fig.width=7}
plot_eigen(eigen_tpdm$vectors, 
           num_vecs = 5, 
           var_names = colnames(financial_data[-1]))
```

This plot shows that the first (and by-far most important) eigenvector
is nearly constant which suggests that large losses in any sector are frequently 
accompanied by large losses across the entire market. The next eigenvalues
distinguish between sub-components of the financial market. The second
eigenvector suggests that the beer, food, health, household, and smoke sectors 
of the market tend to have less extreme losses than more discretionary items 
like autos, books, etc. 

If we believe that the data are tail-equivalaent (i.e., they share an alpha, 
which seems plausible given the plot produced by `alpha_plot()` earlier) then we 
can perform a similar PCA for the covariance analogue (rather than the 
correlation analogue) by setting `trans_marginal = FALSE`. If we want to ensure 
that the alpha associated with the estimated tpd is the same as what we saw in 
the plot above then we can set `fix_alpha = 3.139` or set `k = 300`. 

The default settings for computing the tpd computes the radial components of the 
pseudo-polar decomposition pairwise. This is more sensible if
we expect the most extreme events to be local (e.g., rainfall). When we expect
large events to occur globally it makes sense to compute the radii as a norm 
across the whole vector. This is done with `vector_norm = TRUE`. 

```{r "cov_analogue", fig.height=2.5, fig.width=7}
tpdm_cov <- tpd(as.matrix(-financial_data[,-1]), 
                trans_marginal = FALSE, 
                vector_norm = TRUE)
eigen_cov <- get_eigen(tpdm_cov, make_pd = TRUE)
plot_eigen(eigen_cov$vectors, var_names = colnames(financial_data[-1]))
```

Note the differences between these two sets of eigenvectors (e.g., look at the 
"coal" sector). A primary reason for this is the differences in scale which is 
obscured in the correlation-analogue set of eigenvectors because we fix the 
scales at 1 in that case.

```{r "scale_differences"}
# correlation analogue
head(sort(diag(tpdm), decreasing = TRUE))

# covariance analogue
head(sort(diag(tpdm_cov), decreasing = TRUE))
colnames(financial_data[-1])[order(diag(tpdm_cov))[30:25]]
```



In addition to these plots, we can make time series plots for a few principal 
components. 

```{r 'time_series_plots', fig.height=6, fig.width=7}
dates <- as.Date(as.character(financial_data[,1]), "%Y%m%d")
ts1_plot <- get_ts_plot(eigen_num = 1, 
                        eigen_vecs = eigen_tpdm$vectors, 
                        data = as.matrix(financial_data[,-1]), 
                        dates = dates)
ts2_plot <- get_ts_plot(eigen_num = 2, 
                        eigen_vecs = eigen_tpdm$vectors, 
                        data = as.matrix(financial_data[,-1]), 
                        dates = dates)
ts3_plot <- get_ts_plot(eigen_num = 3, 
                        eigen_vecs = eigen_tpdm$vectors, 
                        data = as.matrix(financial_data[,-1]), 
                        dates = dates)
                  
cowplot::plot_grid(ts1_plot, ts2_plot, ts3_plot, nrow = 3)
```

There is a very large negative value in principal component 1. What day is that?

```{r 'investigate_ts'}
ts1 <- get_pc_ts(eigen_num = 1, 
                 eigen_vecs = eigen_tpdm$vectors, 
                 data = as.matrix(financial_data[,-1]))
ts1_ordered <- order(ts1)
dates[head(ts1_ordered)]
```

The largest loss is Black Monday, October 19, 1987, which is the largest 
single-day U.S. stock market loss. The second largest loss, March 16, 2020, is 
the second Black Monday of the 2020 stock market crash due to COVID-19.

Finally, we can look at scatter plots of the first principal components. 
```{r 'scatter', fig.height=2, fig.width=7}
scatter_df <- data.frame(pc1 = ts1, 
                         pc2 = get_pc_ts(eigen_num = 2, 
                                         eigen_vecs = eigen_tpdm$vectors, 
                                         data = as.matrix(financial_data[,-1])))
max_val <- max(abs(scatter_df))

radii <- apply(scatter_df, 1, function(x){sqrt(sum(x^2))})
scatter_df$large <- as.factor(
  ifelse(radii > quantile(probs = 0.95, radii), 1, 0))

ggplot2::ggplot(scatter_df, ggplot2::aes(x = pc1, y = pc2)) + 
  ggplot2::geom_point(ggplot2::aes(col = large), alpha = 0.5) + 
  ggplot2::coord_fixed() + 
  ggplot2::scale_color_discrete(type = scico::scico(4, palette = "vik")[2:3]) + 
  ggplot2::theme_minimal()
```

In this brief example we demonstrated a few of the key PCA analogue functions: 

* `alpha_plot()` - for checking the assumption of a common alpha.
* `tpd()` - for estimating the TPD matrix of in a multivariate setting.
* `get_eigen()` - for performing the eigen decomposition of a positive definite
    matrix that is close to the estimated TPDM
* `plot_scree()` - for creating a scree plot of the eigenvalues. 
* `plot_eigen()` - for creating a plot which shows eigenvector loadings.  
* `get_ts_plot()` - for creating a time series plot of a principal component. 
* `get_pc_ts()` - for getting the principal component time series for further
    investigation. 
