---
title: "pnw_precip"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tpdmethods)

precip_data <- readRDS("~/Documents/Research/Principle_Component_Analysis/pnw_precip_data_pkg_subset.rds")
dim(precip_data)
precip_data[1:10, 1:6]
```



```{r "transform_margins"}
precip_alpha2 <- apply(precip_data[-c(1,2), -c(1)], 2, 
                       transform_marginal, 
                       use_gpd = FALSE)
lat_lon <- data.frame("lon" = precip_data[1, -c(1)],
                      "lat" = precip_data[2, -c(1)])
dates <- as.Date(as.character(precip_data[-c(1,2), 1]), "%Y%m%d")
```

```{r "tpdm_estimation", cache = TRUE}
tpdm <- matrix(NA, nrow = dim(precip_alpha2)[2], ncol = dim(precip_alpha2)[2])
tictoc::tic("tpdm estimation")
tpdm <- tpd(precip_alpha2, trans_marginal = FALSE)
tictoc::toc()
tictoc::toc()
```

```{r "eigen_decomposition"}
eigen_tpdm <- get_eigen(tpdm, make_pd = TRUE)
```

```{r "eigenvector plots", fig.height=6, fig.width = 6}
eigen_plots <- lapply(1:8, plot_eigen_map, 
                      eigen_vecs = eigen_tpdm$vectors, 
                      eigen_vals = eigen_tpdm$values, 
                      lat = lat_lon$lat, 
                      lon = lat_lon$lon)

cowplot::plot_grid(plotlist = eigen_plots, nrow = 2, ncol = 2, align = "hv")
```

```{r "transformed_eigen_plots", fig.height=6, fig.width=6}
t_eigen_plots <- lapply(1:8, plot_eigen_map, 
                      eigen_vecs = f(eigen_tpdm$vectors), 
                      eigen_vals = eigen_tpdm$values, 
                      lat = lat_lon$lat, 
                      lon = lat_lon$lon)
cowplot::plot_grid(plotlist = t_eigen_plots, nrow = 2, ncol = 2, align = "hv")
```

The eigenvalues indicate that the vast majority of the scale here is 
attributable to the first eigenvector which is an overall storm intensity vector
with few areas of distinction. The remaining eigenvectors appear to split up the
storm into the different regions. Of note is Mt. Rainier in the southern portion
of the map. It is apparent that the mountain regularly sees different rainfall 
than the surrounding regions in large storms. 

Another way to look at the proportional scales of the eigenvalues is to use a 
scree-plot:

```{r "pnw_scree", fig.width=6, fig.height=2.5}
plot_scree(eigen_vals = eigen_tpdm$values)
```

While this may suggest that a single eigenvector is enough, a storm
reconstruction demonstrates that we need several eigenvectors to get the finer
scale features we expect in an extreme storm:


```{r "storm reconstruction", fig.width = 9, fig.height = 5}
storm_totals <- rowSums(precip_alpha2)
max_rain <- which.max(storm_totals) # find the biggest storm

storm_plots <- plot_reconstruction(data = precip_alpha2, 
                                   storm_ind = max_rain, 
                                   dates = dates, 
                                   lat = lat_lon$lat, 
                                   lon = lat_lon$lon, 
                                   eigenvectors = eigen_tpdm$vectors,
                                   eigenvalues = eigen_tpdm$values, 
                                   reconstruction_inds = c(1, 2, 5, 100, 1000))

cowplot::plot_grid(plotlist = storm_plots, nrow = 2, ncol = 3, align = "hv")
```

We note that using five of the `r length(eigen_tpdm$values)` eigenvectors 
gives a pretty good map of the storm but that there are some finer scale 
features that don't show up unless we use many eigenvectors. 


```{r "PC_time_series", fig.height=5, fig.width=9}

ts_plots <- list()

ts_plots <- lapply(c(1:6), get_ts_plot, 
                   eigen_vec = eigen_tpdm$vectors, 
                   dates = dates, 
                   data = precip_alpha2)


cowplot::plot_grid(plotlist = ts_plots, nrow = 2, ncol = 3, align = "hv")
```
We note that these time series plots have missing values as we only downloaded  observations for the winter. 
